String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-29 2023-1-19/sections/cefa.tex
overview

%
%In~\cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a tuple $(R, Q, \Sigma, \delta, q_0, F, \theta)$ where
  \begin{itemize}
  \item $R = \{r_1, \cdots, r_k\}$ is a finite set of registers, 
    \item $Q, \Sigma, q_0, F$ are as in the definition of NFA,
%    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta \subseteq Q \times \Sigma_\varepsilon \times Q \times \Int^R$ is a transition relation, where $\Int^R$ denotes the updates on the values of registers, 
%  set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\alpha: F \rightarrow \Phi(R)$ attaches to each finite state an accepting condition that is specified by an LIA formula.
%    a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
\end{definition}
The definition of CEFA above is slightly different from the definition in~\cite{atva2020} in the sense that accepting conditions on register values
are attached to final states.

@@

The semantics of CEFA is define as follows. Let $\aut = (R, Q, \Sigma, \delta, q_0, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
The \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{b_1} q_1 \cdots q_{m-1}\xrightarrow[\myvec{v_m}]{b_m} q_m$ such that $w = b_1 \cdots b_m$ (some of $b_i$'s may be $\varepsilon$) and $q_{i-1} \xrightarrow[\myvec{v_i}]{b_i} q_i$ for each $i \in [m]$. A run $q_0 \xrightarrow[\myvec{v_1}]{b_1} q_1 \cdots q_{m-1}\xrightarrow[\myvec{v_m}]{b_m} q_m$ is \emph{accepting} if $q_m \in F$ and $\alpha(q_m)[\myvec{\myvec{v'}/\myvec{r}]$ is true, where $\myvec{v'} = \sum \limits_{j \in [m]} \myvec{v_j}}$. Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [m]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. A string $w$ is \emph{accepted} by $\aut$ if there is an accepting run of $\aut$ on $w$. The language defined by $\aut$, denoted by $\Lang(\aut)$, is the set of strings that are accepted by $\aut$.

@@

\zhilin{this example should be modified to match the running example}
\begin{example} \label{eg:1}
  Fig.\ref{fig:repetition} illustrates the CEFA recognizing $(ab)\{1,100\}$. The register vector is $(r_1)$, and the linear arithmetic constraint on $q_2$ is $1\leq r_1\leq 100$. Intuitively, the transition $q_0 \xrightarrow[(1)]{a} q_1$ accept the char $a$ and increase 1 on the value of $r_1$. The transition $q_1 \xrightarrow[(0)]{b} q_2$ accepts the char $b$ and does not change the value. The transition $q_2 \xrightarrow[(0)]{\epsilon} q_2$ is a nondeterministic choice back to initial state $q_0$. Because of the linear arithmetic constraint $1\leq r_1\leq 100$, the value of $r_1$ in the accepting run must be $[1,100]$. $\mathcal{V}(r_1)$ equal to the occurrence number of the transition $q_0 \xrightarrow[(1)]{a} q_1$, so that the accepted string is $ab\cdots ab$ in which $a$ appears at least once and at most 100 times. That is, the language of the automaton is $(ab)\{1,100\}$.
  \begin{figure}[h]
    \centering
    \input{figures/cefa_example.tex}
    \caption{An automaton recognizing the language of $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$}
    \label{fig:repetition}
  \end{figure}
\end{example}

>>>>

%
%In~\cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a tuple $(R, Q, \Sigma, \delta, q_0, F, \theta)$ where
  \begin{itemize}
  \item $R = \{r_1, \cdots, r_k\}$ is a finite set of registers, 
    \item $Q, \Sigma, q_0, F$ are as in the definition of NFA,
%    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta \subseteq Q \times \Sigma \times Q \times \Int^R$ is a transition relation, where $\Int^R$ denotes the updates on the values of registers, 
%  set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\alpha: F \rightarrow \Phi(R)$ attaches to each finite state an accepting condition that is specified by an LIA formula.
%    a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
\end{definition}
The definition of CEFA above is slightly different from the definition in~\cite{atva2020} in the sense that accepting conditions on register values
are attached to final states.

@@

The semantics of CEFA is define as follows. Let $\aut = (R, Q, \Sigma, \delta, q_0, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
The \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$ and $\alpha(q_n)[\myvec{\myvec{v'}/\myvec{r}]$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}}$. Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. A string $w$ is \emph{accepted} by $\aut$ if there is an accepting run of $\aut$ on $w$. The language defined by $\aut$, denoted by $\Lang(\aut)$, is the set of strings that are accepted by $\aut$.

@@

\zhilin{this example should be modified to match the running example}
\begin{example} \label{eg:1}
  Fig.\ref{fig:counting} illustrates the CEFA recognizing $(ab)\{1,100\}$. The register vector is $(r_1)$, and the linear arithmetic constraint on $q_2$ is $1\leq r_1\leq 100$. Intuitively, the transition $q_0 \xrightarrow[(1)]{a} q_1$ accept the char $a$ and increase 1 on the value of $r_1$. The transition $q_1 \xrightarrow[(0)]{b} q_2$ accepts the char $b$ and does not change the value. The transition $q_2 \xrightarrow[(0)]{\epsilon} q_2$ is a nondeterministic choice back to initial state $q_0$. Because of the linear arithmetic constraint $1\leq r_1\leq 100$, the value of $r_1$ in the accepting run must be $[1,100]$. $\mathcal{V}(r_1)$ equal to the occurrence number of the transition $q_0 \xrightarrow[(1)]{a} q_1$, so that the accepted string is $ab\cdots ab$ in which $a$ appears at least once and at most 100 times. That is, the language of the automaton is $(ab)\{1,100\}$.
  \begin{figure}[h]
    \centering
    \input{figures/cefa_example.tex}
    \caption{An automaton recognizing the language of $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$}
    \label{fig:counting}
  \end{figure}
\end{example}

### 2023-04-29 2023-1-19/sections/overview.tex
overview

>>>>

In this section, we utilize the string constraint in Equation~(\ref{eqn-running}) to illustrate the approach in this work.

%$x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120$.

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
Then we first construct a nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$, then add the updates of registers to transitions, resulting into a so-called cost-enriched finite automaton (CEFA) $\aut_1$,  as illustrated in Figure~\ref{fig:overview}(a). 
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $5 \le r_1 \le 60 \wedge 5 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$ that is attached to the final state $p_2$. Note here we represent the counting operators symbolically by registers, instead of unfolding them explicitly. Moreover, we also construct a CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)), which is actually an NFA, since $\Sigma^* c^+$ contains no (explicit) counting operators. 
%where the register $r_0$ is used to record the length of strings.

% the register $r_1$ is incremented in the two transitions from $q_0$ to $q_0$ and $q_1$ respectively,  the register $r_2$ is incremented in the two transitions from $q_1$ to $q_1$ and $q_2$ respectively, and the register $r_3$ is incremented in the transition from $q_2$ to $q_2$ itself. Moreover, the register $r_4$ is incremented in each transition.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.9\textwidth]{sections/overview-cefa.pdf}
  \caption{CEFA for $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$ and $\Sigma^* c^+$}
  \label{fig:overview}
\end{figure}

Next, 
%we remove the $\varepsilon$-transitions of $\aut_1$, resulting into $\aut'_1$, 
we construct $\aut_1 \times \aut_2$, that is, the product of $\aut_1$ and $\aut_2$, as illustrated in Figure~\ref{fig:overview:product}(a). Furthermore, we add one special register, say $r_0$, to record the length of strings. Note that $r_0$ is incremented in all transitions. In $\aut_1 \times \aut_2$, the accepting condition $5 \le r_1 \le 60 \wedge 5 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_0 > 120$ is attached to both $(p_0,q_1)$ and $(p_1,q_1)$, where $r_0 > 120$ is added as a conjunct to express $|x| > 120$.  For technical convenience, we also think the updates on the values of registers in transitions as vectors $(v_1, v_2, v_3, v_4)$, where $v_i \in \Int$ is the update on the register $r_{i-1}$ for each $i \in [4]$. For instance, the transitions corresponding to the self-loop around $(p_0, q_0)$ are thought as $(q, a', q', (1,1,0,0))$ with $a' \in \Sigma \setminus \{a\}$, since $r_0$ and $r_1$ are incremented in these transitions. With the updates thought as vectors, the resulting CEFA is illustrated in Figure~\ref{fig:overview:product}(b).
%the transitions corresponding to the self-loop is $(q, a', q', (1,1,0,0))$ with $a' \in \Sigma \setminus \{a\}$.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.9\textwidth]{sections/overview-cefa-product.pdf}
  \caption{Product of $\aut_1$ and $\aut_2$, where the length register $r_0$ is added}
  \label{fig:overview:product}
\end{figure}

Let $\tau_1, \cdots, \tau_m$ be an enumeration of the transitions of $\aut_1 \times \aut_2$ and $\myvec{v_1}, \cdots, \myvec{v_m}$ be the vectors of updates on the register values in these transitions. For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
Then the satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \times \aut_2$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi(i_1, \cdots, i_m)$ can be computed to express the constraints on the number of occurrences of $\tau_1, \cdots, \tau_m$ in the accepting runs of $\aut_1 \times \aut_2$, where $i_1, \cdots, i_m$ are the integer variables representing the number of occurrences of $\tau_1, \cdots, \tau_m$ respectively. 
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA}
\psi(i_1, \cdots, i_m) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [m]}  i_k v_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
\end{equation}
where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut_1 \times \aut_2$. Finally, the satisfiability of LIA formula can be solved by the off-the-shelf SMT solvers for LIA.

Nevertheless, since it is well known that the satisfiability of existential LIA formulas is NP-complete (see e.g. \cite{Haase18}), the performance could be improved if the size of the formula in~(\ref{eqn-LIA}) can be reduced. Since the size of the formula in~(\ref{eqn-LIA}) depends on the size of $\aut_1 \times \aut_2$, that is, the number of states, transitions, and registers in $\aut_1 \times \aut_2$, we propose the techniques to reduce the size of the CEFA $\aut_1 \times \aut_2$.

Our main idea to reduce the size of the CEFA $\aut_1 \times \aut_2$ is to \emph{ignore the characters in transitions}, since the updates of the values of registers in transitions are independent of the characters therein and the satisfiability of the formula in~(\ref{eqn-LIA}) only concerns about the values of registers. After the characters in transitions are ignored, each transition in $\aut_1 \times \aut_2$ is of the form $(q, \myvec{v}, q')$, where $\myvec{v}$ is the vector denoting the updates on the values of registers. Let $\cB$ denote the resulting automaton. We then think $\cB$ as an NFA, where the vectors $\myvec{v}$ are taken as the characters. Then we apply the determinization and minimization algorithm of NFAs to $\cB$, resulting into an automaton $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.5\textwidth]{sections/overview-cefa-reduced.pdf}
  \caption{Reduced automaton $\cC$ after ignoring the characters in $\aut_1 \times \aut_2$ and applying determinization and minimization}
  \label{fig:overview:product:reduced}
\end{figure}

\zhilin{stopped here}

\hide{
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}
}

### 2023-04-29 2023-1-19/sections/preliminaries.tex
CEFA
overview

\paragraph*{Finite state automata.} 
A \emph{(nondeterministic) finite state automaton} (NFA)  is a tuple $\NFA=(Q, \Sigma, \delta, q_0, F)$, where $Q$ is a finite set of states, $\Sigma$ is a finite alphabet, $\delta \subseteq Q \times \Sigma \times Q$ is the transition relation, $q_0 \in Q$ is the initial state, and $F \subseteq Q$ is the set of final states. For readability, we write a transition $(q, a, q') \in \delta$ as $q \xrightarrow[\delta]{a} q'$ (or simply $q \xrightarrow{a} q'$). %Moreover, when $\delta$ is clear from context, we omit $\delta$ in $q \xrightarrow[\delta]{a} q'$ and write $q \xrightarrow{a} q'$. 
The \emph{size} of an NFA $\NFA$, denoted by $|\NFA|$, is defined as the number of transitions of $\NFA$.
%
A \emph{run} of $\NFA$ on a string $w = a_1 \cdots a_n$ is a sequence of transitions $q_0 \xrightarrow{a_1} q_1 \cdots q_{n-1} \xrightarrow{a_n} q_n$. The run is \emph{accepting} if $q_n \in F$.
A string $w$ is accepted by an NFA $\NFA$ if there is an accepting run of $\NFA$ on $w$. In particular, the empty string $\varepsilon$ is accepted by $\NFA$ if $I \cap F \neq \emptyset$. The language of $\NFA$, denoted by $\Lang(\NFA)$, is the set of strings accepted by $\NFA$. 
%
An NFA $\NFA$ is said to be \emph{deterministic} if $I$ is a singleton and, for every $q \in Q$ and $a \in \Sigma$, there is at most one state $q' \in Q$ such that $(q, a, q') \in \delta$.
%
It is well-known that finite automata capture regular languages precisely. Moreover, the class of regular languages is closed under union, intersection, Kleene star, complement, and language difference \cite{HU79}.

@@

We will also use standard  quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulae, which are typically ranged over by $\phi, \varphi$, etc. %, which is essentially Presburger

>>>>

\paragraph*{Finite automata.} 
A \emph{(nondeterministic) finite automaton} (NFA)  is a tuple $\NFA=(Q, \Sigma, \delta, q_0, F)$, where $Q$ is a finite set of states, $\Sigma$ is a finite alphabet, $\delta \subseteq Q \times \Sigma \times Q$ is the transition relation, $q_0 \in Q$ is the initial state, and $F \subseteq Q$ is the set of final states. For readability, we write a transition $(q, a, q') \in \delta$ as $q \xrightarrow[\delta]{a} q'$ (or simply $q \xrightarrow{a} q'$). %Moreover, when $\delta$ is clear from context, we omit $\delta$ in $q \xrightarrow[\delta]{a} q'$ and write $q \xrightarrow{a} q'$. 
The \emph{size} of an NFA $\NFA$, denoted by $|\NFA|$, is defined as the number of transitions of $\NFA$.
%
A \emph{run} of $\NFA$ on a string $w = a_1 \cdots a_n$ is a sequence of transitions $q_0 \xrightarrow{a_1} q_1 \cdots q_{n-1} \xrightarrow{a_n} q_n$. The run is \emph{accepting} if $q_n \in F$.
A string $w$ is accepted by an NFA $\NFA$ if there is an accepting run of $\NFA$ on $w$. In particular, the empty string $\varepsilon$ is accepted by $\NFA$ if $I \cap F \neq \emptyset$. The language of $\NFA$, denoted by $\Lang(\NFA)$, is the set of strings accepted by $\NFA$. 
%
An NFA $\NFA$ is said to be \emph{deterministic} if $I$ is a singleton and, for every $q \in Q$ and $a \in \Sigma$, there is at most one state $q' \in Q$ such that $(q, a, q') \in \delta$.
%
It is well-known that finite automata capture regular languages precisely. Moreover, the class of regular languages is closed under union, intersection, Kleene star, complement, and language difference \cite{HU79}.

@@

We will also use standard  quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulae, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $X$ of variables, we use $\Phi(X)$ to denote the set of quantifier-free LIA formulas whose free variables are from $X$.
%, which is essentially Presburger
